---
title: "Intro"
format: html
editor: visual
---

## Gapminder Data

We are going to use some of the data on https://www.gapminder.org for this workshop.

Their statement says: **`Our mission is to fight devastating ignorance with a fact-based worldview everyone can understand.`**

## Load packages and import data

```{r}
library('dplyr')
library('tidyr')
library('here')
library('ggplot2')
library('plotly')
```

We have already imposed some structure on the contents of this repository. The `data` folder has three sub-folders -- `raw`, `intermediate`, and `processed`. We can try to import the `raw` data as follows:

```{r}
#| echo: true
#| code-overflow: wrap
#| tidy.opts: { width.cutoff: 60 }
gm_lex_raw <- read.csv(
  '~/reproducibility_r/data/raw/GM_lex_by_country.csv')
```

This works because we are in a container and using relative paths. We can also use the here package to build up the path:

```{r}
here::here()
```

We can add the rest of the path as strings:

```{r}
here::here("reproducibility_r", "data", "raw", "GM_lex_by_country.csv")
```

Or as a string of the full path

```{r}
here::here("reproducibility_r/data/raw/GM_lex_by_country.csv")
```
