---
title: "analysis.Rmd"
output: html_document
date: "2023-09-05"
---



# Load Packages -- touch on .libPaths(), approx 5 minutes
```{r}

```

# Import Data using some hardcoded full paths, then clean it up to be relative path, also discuss how this might work using the `here` package, setwd and getwd, approx 5 minutes, maybe talk about list.files as well.

```{r}

```

# Name some variables some silly things and then talk about why they are silly, talk about variable naming conventions

```{r}

```

# Do some analysis that is badly formatted (e.g., very long nested functions inside functions and stuff)
```{r}

```

# Do some analysis where I iterate over every element of a list, then turn it into an lapply
```{r}

```


# Do some analysis that could be a function
```{r}

```

# Explain what a function is, the parts, how to run a function. Pause for 15 minutes to give people time to make their own functions... 
```{r}

```

# Regroup, discuss what people did, talk about adding some other complexity to your functions (like, adding some sort of 'test' to confirm that the input is numeric or whatever -- see https://swcarpentry.github.io/r-novice-gapminder/10-functions.html) -- might need to add some extra text to talk about data types?

```{r}

```

# Talk about if_else statements (https://swcarpentry.github.io/r-novice-gapminder/07-control-flow.html), give some short activity re: using if_else statements to make some figures or tables or something
```{r}

```

# Talk about annotating and organizing ggplot figures with patchwork (https://patchwork.data-imaginist.com/articles/guides/annotation.html)

```{r}

```

# Spend a few minutes discussing variable scopes
```{r}

```

# Try putting function in a separate .R file and importing that function into the R notebook
```{r}

```

# Talk about bookdown and knitr and quarto  -- part of making code reproducible is combining the analysis and results -- show some examples of how they could separate the notebook into chapters and knit/render it

```{r}

```

# Talk about importing data from PDF files w pdftools
```{r}

```

# Talk about reading and exporting excel files -- is there a way to do conditional formatting? ppl love it for reasons
```{r}

```

# Maybe talk about knitting nice tables too?
```{r}

```

# Interactive stuff with ggplotly, caution about some of the default settings in ggplotly being a little different
```{r}

```

