---
title: "Data verification with `assertr`"
format: 
  html:
    self-contained: true
author: Joselynn Wallace & George Dang
---

# assertr
Another package that could help with making your data analysis more reproducible is `assertr` (https://cran.r-project.org/web/packages/assertr/assertr.pdf). It is basically a suite of functions to check that your data meets some of your assumptions before you proceed with your analysis. 

```{r}
library(magrittr)
library(assertr)
```

Let's work with the mtcars data. The mpg column is an example of data that shouldn't have any negative values:
```{r}
head(mtcars)
mtcars %>%
  dplyr::group_by(cyl) %>%
  dplyr::summarise(avg.mpg=mean(mpg))
```

What if this wasn't true? We can mess with the data a little to see what happens..

```{r}
mtcars$mpg[5] 
mtcars$mpg[5] <- mtcars$mpg[5] * -1
```

Then we can use `assertr::verify` to add a step where we confirm that we don't have any negative values first, and if we do `assertr` will stop the process
```{r}
mtcars %>%
  assertr::verify(mpg >= 0) %>%
  dplyr::group_by(cyl) %>%
  dplyr::summarise(avg.mpg=mean(mpg))
```

