---
title: "Reproducibility in R"
format: 
  html:
    self-contained: true
author: Joselynn Wallace & George Dang
---

# Here

When looking at someone elses R code, you might see scripts or notebooks with lines with hard-coded paths, like this:

```{r, eval=F, echo=T}
data <- read.table('/Users/Joselynn/Desktop/raw_data.txt')
```

Or maybe start like this:
```{r, eval=F, echo=T}
setwd('/Users/Joselynn/Desktop')
```

These approaches aren't *wrong*, they are a bit fragile and could break when your colleagues try to run your code on their computers. One way to avoid this is to use RStudio projects and the `here` package. RStudio projects are nice because they let you keep separate working environments for different projects and do lots of other very nice things for you (https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects). You can make new RStudio projects through the RStudio app or with the `usethis` package (https://usethis.r-lib.org/reference/index.html). I've already set this repository up as an RStudio project. You can open the project by going up to `File > Open Project` and opening the `reproducibility_r.Rproj` in the base of this repo. Now you can use `here` to refer to paths:

```{r}
here::here()
```

And you can build up full paths like this:

```{r}
list.files(here::here('notebooks/book/_book'))
```

You can `setwd` to change the working directory and `here::here` should still work:
```{r}
setwd('~')
getwd()
here::here()
```
